\graphicspath{{content/chapters/3_literature_review/figures/}}

\chapter{Literature Review}%
\label{chp:literature_review}

\section{AI in Language Learning}
\label{sec:ai_in_language_learning}

\section{Question Answering Systems}
\label{sec:question_answering_systems}

Knowledge-based Question Answering

\section{Chatbots \& Dialogue Systems}
\label{sec:chatbots_dialogue_systems}

\section{Datasets}
\label{sec:datasets}

\subsection{Universal Morphology}
\label{sec:universal_morphology}

\gls{unimorph}

unimorph~\cite{unimorph}

\subsection{Treebanks}
\label{sec:treebanks}

PROIEL Treebank~\cite{proiel}, Perseus Latin Dependency Treebank~\cite{10.1007/978-3-642-20227-8_5}

\section{Tools and Digital Resources for Latin}
\label{sec:tools_and_digital_resources_for_latin}

William Whitaker's Words, \gls{cltk}

\section{BERT and Latin BERT}
\label{sec:bert_and_latin_bert}

% \gls{bert} , Latin \gls{bert}~\cite{latinbert}



\gls{bert}, introduced by~\cite{devlin-etal-2019-bert},
is a language representation model that implements
a Transformer encoder architecture that processes text 
bidirectionally by considering both left and right context
simultaneously.

\gls{bert} is pre-trained on unlabeled data using two unsupervised 
tasks, a masked language model and next sentence prediction. 
This pre-training enables \gls{bert} to learn general language 
representations. The pre-trained \gls{bert} model can then be 
fine-tuned for downstream tasks, such as intent classification, 
by adding task-specific layers and training on labelled data 
for that task. This transfer learning technique is very 
effective, particularly for low-resource tasks, because 
the model makes use of knowledge from pre-training, rather 
than having to learn entirely from scratch~\cite{devlin-etal-2019-bert}. 
For this project, fine-tuning \gls{bert} for the intent 
classification task on Latin grammar questions will enable 
the model to recognize certain patterns such as question 
structures and grammar-related terminology.

% Latin BERT

Recognizing the limitations of generic pre-trained models 
for Latin-specific tasks,~\cite{latinbert} developed 
Latin \gls{bert}, a \gls{bert} model trained specifically on Latin texts. 
Latin \gls{bert} was trained on a large corpus of Latin text 
from several sources, enabling it to learn Latin-specific 
language representations that generic multilingual \gls{bert} 
models might not capture effectively, which offers many 
improvements for Latin-specific \gls{nlp} tasks. 
\cite{latinbert} demonstrate the usefulness of Latin 
\gls{bert} with four case studies of different \gls{nlp} tasks, 
showing Latin \gls{bert} outperforms other models. First, 
they show that Latin \gls{bert} produces a new state of the 
art for POS tagging on three different datasets. Then they 
show Latin \gls{bert}'s ability to be used for the task of text 
infilling. It also outperforms static embeddings for the 
task of word sense disambiguation. Finally, they demonstrate 
Latin \gls{bert}'s potential for contextual nearest neighbours.

\section{Intent Classification}
\label{sec:intent_classification}

% TODO: FIND CITATIONS FOR THIS SECTION

Intent classification is an important component of 
certain dialogue systems, enabling chatbots to 
understand the user's goals based on natural language input. 
It is commonly used in customer service chatbots, 
task-oriented dialogue systems, and question-answering 
systems. Intent classification involves categorizing 
user utterances according to their underlying goal. 
The dialogue system uses the classified intent to determine 
what action to take. Accurate intent classification is 
very important to chatbot effectiveness because 
misclassification leads to inappropriate responses.

Various approaches have been applied to intent classification 
as listed by~\cite{hoque2025intent} and~\cite{10.1145/3718491.3718520}. 
Traditional intent classification approaches have used 
machine learning methods such as \gls{svm}, 
logistic regression, and Naive Bayes. 
Deep learning based approaches such as \gls{cnn}
and \gls{rnn} have also been used for this task. 
More recently, the use of transformer-based models, 
such as \gls{bert}, for intent classification has achieved good 
results. For instance,~\cite{chen2019bertjointintentclassification} 
propose a joint intent classification and slot filling model 
based on \gls{bert}, with results showing that it outperformed 
attention-based \gls{rnn} models and slot-gated models on benchmark 
datasets.

This project adapts intent classification to the specific 
domain of Latin grammar by treating grammatical concepts as 
intents in a dialogue system. This adaptation reframes the 
intent classification task as identifying which grammatical 
aspect the learner is asking about.

\section{Morphological Analysis}
\label{sec:morphological_analysis}

% about morphological analysis itself

% morphological analysis for the latin language


